---
title: "Nobel Prize Winner's Relative Ages"
author: "Maja Založnik "
date: "compiled `r Sys.time()`"
output: html_document
---
# Introduction
This document uses the Nobel Prize API data on Nobel Prize winners' ages and the Our world in Data data on life expectancies, to calculate the "relative" ages of the winners, and supplements the Oxford Institute of Population Ageing blog post of 12.10.2016

HT to Neil Saunders who's [blog post](https://nsaunders.wordpress.com/2016/10/09/lets-briefly-revisit-the-nobel-api/) alerted me to the Nobel API and prompted this back of the envelope calculatio. His very useful code for analysing the API data is in  [this github repo here](https://github.com/neilfws/NobelAnalysis), and I stole from it liberally - all additional kludginess is purely my own. 


# Getting the data


```{r setup, message=FALSE}
# preliminaries
library(jsonlite)
library(dplyr)
```

## Nobel prize laureates - and some errors?
We use the [Nobel Prize API](https://nobelprize.readme.io/) to fetch laureate data in JSON format. (at one point I got really confused (violenlty so) with the nested data.frames produced by fromJSON, and found this [post](https://cran.r-project.org/web/packages/tidyjson/vignettes/introduction-to-tidyjson.html) on tidyjson which looks great, TODO!)

```{r get-data, tidy=TRUE, message=FALSE}
nobel <- fromJSON("http://api.nobelprize.org/v1/laureate.json")
```

Top level is `nrow(nobel$laureates)= ``r nrow(nobel$laureates)` laureates. Six of which have more than one prize, which you can tell by looking at how many rows their `nobel$laureates$prizes` variables have:

```{r}
cnt <- sapply(nobel$laureates$prizes, function(x) nrow(x))
multi <- which(cnt > 1)
nobel.multi.df <- cbind(bind_rows(nobel$laureates[multi,])[rep(1:length(cnt[multi]),cnt[multi]),1:2],
            bind_rows(nobel$laureates$prizes[multi])[1:2])
nobel.multi.df
```

That's all good and well, but when you do the whole table for all the prizes, it turns out that there are in fact six entries in the original table that should not be there. They are empty rows - but for the gender of these laureates being recorded as male. 

```{r}
nobel$laureates[is.na(nobel$laureates$firstname),]
```

So that's not good, these need to be removed and someone possibly needs to be alerted about this (TODO: open issue with neilfws, he's got the same error; also have a look at Noble API contacts..)

OK, so now we want our dataframe with a single row per prize, with the recipients' names, DOB, year awarded and category. 

```{r}
#merge into final df
nobel.df <- cbind(bind_rows(nobel$laureates)[rep(1:length(cnt),cnt),c(1:4,12)],
            bind_rows(nobel$laureates$prizes)[1:2])
# remove weird six rows
nobel.df <- nobel.df[!is.na(nobel.df$firstname),]

```

Now as od 11.10. this data was not teriffically up to date, since the new 10 recipients' birthdays were not picked up yet, although their names etc. had been. So these need to be added manually, but also in a way that if this is rerun in the future it doesn't overwrite actual data pulled from the API. 


```{r}
# add new 2016 one's birthdays 
# https://en.wikipedia.org/wiki/Yoshinori_Ohsumi
# http://starsunfolded.com/michael-kosterlitz/ (only one not precise, assume bday before 10.10.)
# https://en.wikipedia.org/wiki/Duncan_Haldane
# https://en.wikipedia.org/wiki/David_J._Thouless
# https://en.wikipedia.org/wiki/Fraser_Stoddart
# https://en.wikipedia.org/wiki/Jean-Pierre_Sauvage
# https://en.wikipedia.org/wiki/Ben_Feringa
# https://en.wikipedia.org/wiki/Juan_Manuel_Santos
# https://en.wikipedia.org/wiki/Oliver_Hart_(economist)
# https://en.wikipedia.org/wiki/Bengt_R._Holmstr%C3%B6m

nobel.df %>% 
  mutate(born = ifelse(surname == "Ohsumi" & born =="0000-00-00","1945-02-09", born)) %>%
  mutate(born = ifelse(surname == "Kosterlitz" & born =="0000-00-00", "1942-01-01", born)) %>%
  mutate(born = ifelse(surname == "Haldane" & born =="0000-00-00", "1951-09-14", born)) %>%
  mutate(born = ifelse(surname == "Thouless" & born =="0000-00-00", "1934-09-21", born)) %>%
  mutate(born = ifelse(surname == "Stoddart" & born =="0000-00-00", "1942-05-24", born)) %>%
  mutate(born = ifelse(surname == "Sauvage" & born =="0000-00-00", "1944-10-21", born)) %>%
  mutate(born = ifelse(surname == "Feringa" & born =="0000-00-00", "1951-05-18", born)) %>%
  mutate(born = ifelse(surname == "Santos" & born =="0000-00-00", "1951-08-10", born)) %>%
  mutate(born = ifelse(surname == "Hart" & born =="0000-00-00", "1948-10-09", born)) %>%
  mutate(born = ifelse(surname == "Holmström" & born =="0000-00-00", "1949-04-18", born)) ->
  nobel.df

```


Now just need to calculate the age at the time of being awarded the prize, and remove all the organizations, keeping humans only
```{r}
nobel.df %>%
  mutate(age = as.numeric(as.Date(paste(year, "12-31", sep = "-"), 
                                  "%Y-%m-%d") - as.Date(born,"%Y-%m-%d"))/365,
         year = as.numeric(year)) %>%
  filter(gender != "org") -> nobel.df
           
rm(cnt, nobel, nobel.multi.df)
```

## Life expectancy

An earlier version had me parsing the whole Human Mortalitdy Database Ex data, but not only would the country mathcing be a pain, it also makes sense conceptually to look at the gloal life expectancy when measuring the relative age of mankinds biggest contributors :) 

But HMD doesn't have world data, Max does. The data points are a bit sparse early on.. I use zoo spline to interpolate the missing values. 
```{r, echo = FALSE}
lexp <- read.csv("https://ourworldindata.org/grapher/life-expectancy-globally-since-1770.csv?country=ALL")
#rename stupid column name
names(lexp)[3] <- "lexp"
# keep only world data on year and lexp
# pad missing years and join back
# spline for interpolated lexp

lexp %>% 
  filter(Country == "World") %>%
  select(Year, lexp ) %>%
  full_join(data.frame(Year = seq(min(lexp$Year), 2016))) %>%
  arrange(Year) %>%
  mutate(lexp.interpolated = zoo::na.spline(lexp)) %>%
  select(Year, lexp.interpolated) %>%
  right_join(nobel.df, by = c("Year" = "year"))-> full.lexp

 rm(nobel.df, lexp)

```


And there we have it,  `full.lexp` is the full table of 884 instances of humans being awarded a Nobel peace prise, with their age at award and the (estimated) average global (period) life expectancy at the time of award. 

# Plotting the trends

First just the plotting function and separate the data into little tables, one for each category. 


```{r}
# you can adjust span and degree for the smoothing curve, defaults are 2/3 and 1
span = 2/3
degree = 1
FunPlot <- function(data, color, norm = TRUE){
  x <- data$Year
  y <- data$age
  y.grids <- seq(0,100,20)
  ylim <- 100
  if (!norm) {y.grids <- y.grids/40
  y <- data$age/data$lexp.interpolated
  ylim <- 2.3
  }
  scatter.smooth(x, y, pch = 19, 
                 axes = FALSE, span = span, degree= degree,
                 col = color[1], lpars = list(col = color[2], lwd = 3),
                 xlab = "Year", ylab = "Age", main = deparse(substitute(data)),
                 ylim = c(0,ylim), xlim = c(1900, 2020))
  for (i in y.grids){
      lines(c(1900,2020), c(i,i), lty=2, lwd=0.5)
  }
axis(1, las=2, at = seq(1900, 2020, 20),labels = c(1900, 1920, 1940, 1960, 1980, 2000,""))
}

# and neat little tables for each category
physics <- full.lexp[full.lexp$category == "physics",]
chemistry <- full.lexp[full.lexp$category == "chemistry",]
medicine <- full.lexp[full.lexp$category == "medicine",]
economics <- full.lexp[full.lexp$category == "economics",]
literature <- full.lexp[full.lexp$category == "literature",]
peace <- full.lexp[full.lexp$category == "peace",]

```

### Age of Nobel Prize Winners at the time of award - by category

OK, so the first plot is essentially recreating the one by Neil, the Economist and others..
```{r}

layout(mat=matrix(1:7, c(1,7)), widths = c(1,3,3,3,3,3,3))
par(mar = c(3,2.2,1,0))
plot(1,1, ylim= c(0,100), type = "n", axes = FALSE, xlab = "", ylab = "")
axis(2, las=2)
par(mar = c(3,0,1,0))
FunPlot(physics, c("darkolivegreen3", "darkolivegreen4"))   
FunPlot(chemistry, c("plum2", "plum4"))    
FunPlot(medicine, c("skyblue", "skyblue3"))   
FunPlot(economics, c("orange1", "orange3"))  
FunPlot(literature, c("gray70", "gray40"))  
FunPlot(peace, c("mediumpurple1", "mediumpurple4"))  

```


### Relative Age of Nobel Prize Winners at the time of award - by category - relative to global average life expectancy. 

And now the second set of plots, where the age is divided by the global average period life expectancy at teh time. And there you have it. 

```{r}
layout(mat=matrix(1:7, c(1,7)), widths = c(1,3,3,3,3,3,3))
par(mar = c(3,2.2,1,0))
plot(1,1, ylim= c(0,2.3), type = "n", axes = FALSE, xlab = "", ylab = "")
axis(2, las=2)
par(mar = c(3,0,1,0))
FunPlot(physics, c("darkolivegreen3", "darkolivegreen4"), norm = FALSE)   
FunPlot(chemistry, c("plum2", "plum4"), norm = FALSE)    
FunPlot(medicine, c("skyblue", "skyblue3"), norm = FALSE)   
FunPlot(economics, c("orange1", "orange3"), norm = FALSE)  
FunPlot(literature, c("gray70", "gray40"), norm = FALSE)  
FunPlot(peace, c("mediumpurple1", "mediumpurple4"), norm = FALSE)  

```
               