---
title: "Nobel Prize Winner's Relative Ages"
author: "Maja Zaloznik"
date: "compiled `r Sys.time()`"
output: 
  html_document: 
  keep_md: yes
number_sections: yes
---

# Introduction
This document uses the Nobel Prize API data on Nobel Prize winners' ages and the Human Mortality Database data on life expectancies, to calculate the "relative" ages of the winners. 

HT to Neil Saunders who's code for analysing the API data is in  [this github repo here](https://github.com/neilfws/NobelAnalysis).


# Getting the data
We use the [Nobel Prize API](https://nobelprize.readme.io/) to fetch laureate data in JSON format.

```{r setup, message=FALSE}
library(jsonlite)
library(dplyr)
require(stringr)
```

```{r get-data, tidy=TRUE, message=FALSE}
u <- "http://api.nobelprize.org/v1/laureate.json"
nobel <- fromJSON(u)

# table with one row per prize
prizes <- rbind_all(nobel$laureates$prizes)[1:2]


# count multiple winners
cnt <- sapply(nobel$laureates$prizes, function(x) nrow(x))
prizes$gender <- rep(nobel$laureates$gender, cnt)
prizes$born <- rep(nobel$laureates$born, cnt)

# add new 2016 one's birthdays 
# 907 Yoshinori Ohsumi medicine February 9, 1945 https://en.wikipedia.org/wiki/Yoshinori_Ohsumi
prizes$born[907] <- "1945-02-09"
# http://starsunfolded.com/michael-kosterlitz/
prizes$born[908] <- "1942-01-01"
# https://en.wikipedia.org/wiki/Duncan_Haldane
prizes$born[909] <- "1951-09-14"
# https://en.wikipedia.org/wiki/David_J._Thouless
prizes$born[910] <- "1934-09-21"
#https://en.wikipedia.org/wiki/Fraser_Stoddart
prizes$born[911] <- "1942-05-24"
#https://en.wikipedia.org/wiki/Jean-Pierre_Sauvage
prizes$born[912] <- "1944-10-21"
#https://en.wikipedia.org/wiki/Ben_Feringa
prizes$born[913] <- "1951-05-18"
#https://en.wikipedia.org/wiki/Juan_Manuel_Santos
prizes$born[914] <- "1951-08-10"
#https://en.wikipedia.org/wiki/Oliver_Hart_(economist)
prizes$born[915] <- "1948-10-09"
#https://en.wikipedia.org/wiki/Bengt_R._Holmstr%C3%B6m
prizes$born[916] <- "1949-04-18"

prizes$age <- as.numeric(as.Date(paste(prizes$year, "12-31", sep = "-"), "%Y-%m-%d") - as.Date(prizes$born, "%Y-%m-%d"))/365
prizes$country <- rep(nobel$laureates$bornCountryCode, cnt)

#for some reason there are 6 empty rows in the data. remove
prizes <- prizes[!is.na(prizes$year),]
prizes$year <- as.numeric(prizes$year)
# remove organizations as well
prizes %>%
  filter(gender != "org") -> prizes

```


Cleaning up life expectancy data from HMD (cheet sheet John Minton's code [here](https://github.com/JonMinton/Lexis_Surface_Difference_Maps) ). 

```{r, eval = FALSE, echo = FALSE}
files.to.grab <- dir("data/e0_per/E0per/")

Fun <- function(filename, dir){
  out <- read.table(
    paste0(
      dir, filename
      ),
    skip=1, header=T
    )
  
  country_code_part <- str_extract(filename, "^[^\\.]*") 
  # regex: 
  # ^ : start of line
  # [^\\.]*  : anything but '.', 1 or more times
  
  out$country_code <- country_code_part

  return(out)
}

# yup, solved it without plyr! (John used ldply)
lexp <- files.to.grab %>% lapply(Fun, dir="data/e0_per/E0per/") %>% rbind_all()

```

Actually, it's easier to use world data, from "our world in data"
```{r}
lexp <- read.csv("https://ourworldindata.org/grapher/life-expectancy-globally-since-1770.csv?country=ALL")
#rename stupid column name
names(lexp)[3] <- "lexp"
# keep only world data on year and lexp
# pad missing years and join back
lexp %>% 
  filter(Country == "World") %>%
  select(Year, lexp ) %>%
  full_join(data.frame(Year = seq(min(lexp$Year), 2016))) %>%
  arrange(Year) %>%
  mutate(lexp.interpolated = zoo::na.spline(lexp)) %>%
  select(Year, lexp.interpolated) %>%
  right_join(prizes, by = c("Year" = "year"))-> full.lexp

 

```


and there we have it! 

```{r}

```
Yoshinori Ohsumi medicine February 9, 1945 https://en.wikipedia.org/wiki/Yoshinori_Ohsumi